Rates of true polar wander in convecting planets

### 2015-08-06 tpw_rate.tex
Moar text

The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through Equation \eqref{eq:grow_perturbation} and its relaxation through Equation \eqref{eq:simple_milankovitch}.
We may plug in the estimate for $\Lambda_{ij}$ into equation \eqref{eq:simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{\mathrm{Fr}} \sin{2 \theta}
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
At high Rayleigh numbers response timescales for relaxation of the mismatch angle go down.
At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d/R \sim \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\dot{\theta} \propto \mathrm{Ra}^{1-3\beta}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.

More important is the observation that the most important parameters are the Froude number, which acts as the brakes on the system, and $\theta(\mathrm{Ra})$.
Whereas the rest of the expression has at most a weak dependence on $\mathrm{Ra}$, this is expected to be strongly dependent on it.

Indeed, we can identify two endmember behaviors of Equation \eqref{eq:simplest_milankovitch}.
When convection is not sufficiently chaotic to create a large $\theta$ we are in the regime where the planet's rotation axis closely tracks that of the convective moment.
This is the regime considered in \citet{steinberger1997changes}, \citet{roberts2007cause}, and \citet{zhong2007supercontinent}, and can be considered the ``slow TPW'' regime.

When convection is more chaotic, however, there may be large excursions in $\theta$, which can drastically increase the wander rate.  
If $\theta=90^\circ$, this corresponds to IITPW \citep{kirschvink1997evidence}.  
This, however, is a special case in the large $\theta$, ``fast TPW'' regime.

@@

The theory of TPW response for the case of elastic lithospheres has been developed in, among other places, 
\citet{matsuyama2006rotational}, \citet{creveling2012mechanisms}, and \citet{chan2014time}.
The formalism developed in Section~\ref{sec:scaling} can still be applied to this case, 
though the response to internal variations in the moment of inertia becomes more limited.

\section{Conclusion}
We have developed a framework for approximating the rates of true polar wander in terms of a small number of nondimensional numbers.

>>>>

The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through Equation \eqref{eq:grow_perturbation} and its relaxation through Equation \eqref{eq:simple_milankovitch}.
We may plug in the estimate for $\Lambda_{ij}$ into equation \eqref{eq:simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{\mathrm{Fr}} \sin{2 \theta}
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
At high Rayleigh numbers response timescales for relaxation of the mismatch angle go down.
At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \sim \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\dot{\theta} \propto \mathrm{Ra}^{1-3\beta}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.

We then suggest that the most important parameters are the Froude number, which acts as the brakes on the system, and $\theta(\mathrm{Ra})$.
Whereas the rest of the expression has at most a weak dependence on $\mathrm{Ra}$, this is expected to be strongly dependent on it.

Indeed, we can identify two endmember behaviors of Equation \eqref{eq:simplest_milankovitch}.
When convection is not sufficiently chaotic to create a large $\theta$ we are in the regime where the planet's rotation axis closely tracks that of the convective moment.
This is the regime considered in \citet{steinberger1997changes}, \citet{roberts2007cause}, and \citet{zhong2007supercontinent}, and can be considered the ``slow TPW'' regime.
When convection is more chaotic, however, there may be large excursions in $\theta$, which can drastically increase the wander rate.  
If $\theta=90^\circ$, this corresponds to IITPW \citep{kirschvink1997evidence}.  
This, however, is a special case in the large $\theta$, ``fast TPW'' regime.

As an example, we may consider the early Earth, when the mantle was presumably hotter and less viscous, leading to a 
higher Rayleigh number. We would then predict that convection was more vigorous, leading to 
a less stable $\theta(t)$, and thus more TPW and more frequent ``fast TPW'' events.

@@

The theory of TPW response for the case of elastic lithospheres has been developed in, among other places, 
\citet{matsuyama2006rotational}, \citet{creveling2012mechanisms}, and \citet{chan2014time}.
The formalism developed in Section~\ref{sec:scaling} can still be applied to this case, 
though the response to internal variations in the moment of inertia becomes more limited 
(and potentially richer, as in the oscillitory motions suggested by \citet{creveling2012mechanisms}).

\section{Conclusion}
We have developed a framework for discussing the rates of true polar wander for a convecting planet 
from a perspective of scaling and fluid dynamics.
We identified a small number of dimensionless parameters which describe the system, and showed how they affect the overall dynamics of the system.

The most important parameters are the Rayleigh number and the Froude number.
The Froude number acts as a damper to TPW.
The Rayleigh number is more complicated, since it is a control on both the forcing 
of TPW and the response, which act in opposite directions.
This perspective allows us to consider not only the polar wandering of Phanerozoic Earth,
but also allows us to hypothesize about polar wandering during the Archean and Proterozoic, or 
on other planetary bodies.
